Manuscript Figures for Genome Biology

### 2018-05-28 datasets/RNA-Seq/polyester-simulation.Rmd
Add uninformative covariate sim to yeast and poly

@@

```{r}
#' @param counts is the full count dataset
#' @param paramPolyester is the parameters from running polyester. 
#' Contains information on the mean-variance relationship used to 
#' simulate RNA-Seq counts. 
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
simulateOneSplit <- function(X, counts, paramPolyester, 
                             nDE, sampleSize, bd){

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(counts),
                               pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = res_alt$baseMean, 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth, 
                               pzero = rowSums(counts(dds_alt)==0)/
                                 ncol(counts(dds_alt))  ))

@@

>>>>

resfile_n5_uninfCov <- file.path(resdir, "polyester-results-null5-uninfCov.rds")
resfile_n10_uninfCov <- file.path(resdir, "polyester-results-null10-uninfCov.rds")
resfile_d5_uninfCov <- file.path(resdir, "polyester-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "polyester-results-de10-uninfCov.rds")

@@

```{r}
#' @param counts is the full count dataset
#' @param paramPolyester is the parameters from running polyester. 
#' Contains information on the mean-variance relationship used to 
#' simulate RNA-Seq counts. 
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
simulateOneSplit <- function(X, counts, paramPolyester, 
                             nDE, sampleSize, bd, uninformativeCovariate = FALSE){

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(counts),
                               pval=res_alt$pvalue, SE=res_alt$lfcSE, 
                               ind_covariate = ifelse(uninformativeCovariate, 
                                                      rnorm(nrow(res_alt)),
                                                      res_alt$baseMean), 
                               effect_size=res_alt$log2FoldChange, 
                               test_statistic=res_alt$stat, qvalue=truth, 
                               pzero = rowSums(counts(dds_alt)==0)/
                                 ncol(counts(dds_alt))  ))

@@

# Uninformative Covariate

## Sim N5: Null Comparison 5v5

### Generate a list of SB results objects

if (!file.exists(resfile_n5_uninfCov)){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(null5, file=resfile_n5_uninfCov)
} else {
  null5 <- readRDS(file=resfile_n5_uninfCov)
}
```

### Plot average results over replications

## Sim N10: Null Comparison 10v10

### Generate a list of SB results objects

if (!file.exists(resfile_n10_uninfCov)){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(null10, file=resfile_n10_uninfCov)
} else {
  null10 <- readRDS(file=resfile_n10_uninfCov)
}
```

### Plot average results over replications

## Sim D5: DE Comparison 5v5

### Generate a list of SB results objects

if (!file.exists(resfile_d5_uninfCov)){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                  paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                  bd=bd, uninformativeCovariate = TRUE)
  saveRDS(de5, file=resfile_d5_uninfCov)
} else {
  de5 <- readRDS(file=resfile_d5_uninfCov)
}
```

### Plot average results over replications

## Sim D10: DE Comparison 10v10

### Generate a list of SB results objects

if (!file.exists(resfile_d10_uninfCov)){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, 
                   bd=bd, uninformativeCovariate = TRUE)
  saveRDS(de10, file=resfile_d10_uninfCov)
} else {
  de10 <- readRDS(file=resfile_d10_uninfCov)
}

### Plot average results over replications

## Sample size comparison

```{r}
# uninformative covariate
plotMethodRanks(c(resfile_d5_uninfCov, resfile_d10_uninfCov), 
                colLabels = c("DE 5", "DE 10"), 
                alpha = 0.10, xlab = "Comparison",
                excludeMethods = NULL)
```

### 2018-05-28 datasets/RNA-Seq/yeast-simulation.Rmd
Add uninformative covariate sim to yeast and poly

This is an analysis of the yeast data with 48 biological replicates in each of 
two conditions (analyzed in this [publication](https://www.ncbi.nlm.nih.gov/pubmed/26206307/)). 
We chose this experiment because of the large number of biological replicates, which
will allow us to (1) implement null comparisons on random subsets of samples within
one condition, (2) compare FDR approaches on their ability to discover 'true positive' 
DE genes using subsets of the full dataset, where 'true positive' is defined as a 
gene detected as DE using all 48 samples in each condition, and (3) start with a null
comparison and add in artificial differences to a subset of genes to define
'true positives'.

@@

@@

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd){

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE, 
                               ind_covariate = resTEST$baseMean,
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test))  ))

@@

>>>>

This is an analysis of the yeast data with 48 biological replicates in each of 
two conditions (analyzed in this [publication](https://www.ncbi.nlm.nih.gov/pubmed/26206307/)). 
We chose this experiment because of the large number of biological replicates, which
will allow us to (1) implement null comparisons on random subsets of samples within
one condition, and (2) start with a null comparison and add in artificial differences
to a subset of genes to define 'true positives'.

We will investigate the overall mean expression as an informative and independent 
(under the null hypothesis) covariate. 
We will also investigate the effects of an uninformative covariate by using a 
randomly generated covariate.

@@

resfile_n5_uninfCov <- file.path(resdir, "yeast-results-null5-uninfCov.rds")
resfile_n10_uninfCov <- file.path(resdir, "yeast-results-null10-uninfCov.rds")
resfile_d5_uninfCov <- file.path(resdir, "yeast-results-de5-uninfCov.rds")
resfile_d10_uninfCov <- file.path(resdir, "yeast-results-de10-uninfCov.rds")

@@

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd, uninformativeCovariate = FALSE){

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE, 
                               ind_covariate = ifelse(uninformativeCovariate, 
                                                      rnorm(nrow(resTEST)),
                                                      resTEST$baseMean),
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test))  ))

@@

# Uninformative covariate

Here we repeat the previous four sections, using an uninformative covariate 
instead of the overall mean expression.

## Sim N5: Null Comparison 5v5

### Generate a list of SB results objects

```{r n5-uc, results='hide', message=FALSE}
rseed <- 225
sampleSize <- 5
nDE <- 0

if (!file.exists(resfile_n5_uninfCov)){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   uninformativeCovariate = TRUE)
  saveRDS(null5, file = resfile_n5_uninfCov)
}else{
  null5 <- readRDS(file = resfile_n5_uninfCov)
}
```

### Plot average results over replications

## Sim N10: Null Comparison 10v10

### Generate a list of SB results objects

```{r n10-uc, results='hide', message=FALSE}
rseed <- 837
sampleSize <- 10
nDE <- 0

if (!file.exists(resfile_n10_uninfCov)){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   uninformativeCovariate = TRUE)
  saveRDS(null10, file = resfile_n10_uninfCov)
}else{
  null10 <- readRDS(file = resfile_n10_uninfCov)
}
```

### Plot average results over replications

## Sim D5: DE Comparison 5v5

### Generate a list of SB results objects

```{r d5-uc, results='hide', message=FALSE}
rseed <- 198
sampleSize <- 5
nDE <- 500

if (!file.exists(resfile_d5_uninfCov)){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   uninformativeCovariate = TRUE)
  saveRDS(de5, file = resfile_d5_uninfCov)
}else{
  de5 <- readRDS(file = resfile_d5_uninfCov)
}
```

### Plot average results over replications

## Sim D10: DE Comparison 10v10

### Generate a list of SB results objects

```{r d10-uc, results='hide', message=FALSE}
rseed <- 961
sampleSize <- 10
nDE <- 500

if(!file.exists(resfile_d10_uninfCov)){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   uninformativeCovariate = TRUE)
  saveRDS(de10, file = resfile_d10_uninfCov)
}else{
  de10 <- readRDS(file = resfile_d10_uninfCov)
}
```

### Plot average results over replications

## Sample size comparison

```{r}
# uninformative covariate
plotMethodRanks(c(resfile_d5_uninfCov, resfile_d10_uninfCov), 
                colLabels = c("DE 5", "DE 10"), 
                alpha = 0.10, xlab = "Comparison", 
                excludeMethods = NULL)
```

### 2018-05-28 manuscript_figures/manuscript_figures.Rmd
Add uninformative covariate sim to yeast and poly

```{r Figure2}
objects <- list.files( path, recursive=TRUE, pattern="rds", full.names=TRUE )
resfile <- objects[grepl("yeast", objects) & grepl("de5", objects)] 
de5 <- readRDS(file=resfile)
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

@@

# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|null|polyester"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|null"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

# null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|de"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|yeast|null"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

>>>>

```{r Figure2}
objects <- list.files( path, recursive=TRUE, pattern="rds", full.names=TRUE )
resfile <- objects[grepl("yeast", objects) & 
                   grepl("de5", objects) & 
                  !grepl("_uninfCov", objects)] 
de5 <- readRDS(file=resfile)
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

@@

# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|null|polyester"), dataset) &
            !grepl("scRNAseq", casestudy) &
            !grepl("_uninfCov", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|null|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

# null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|de|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|yeast|null|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))
